Timeliness of security updates for Android

### 2014-11-10 securityupdates.tex
New introduction

\section{Introduction}
Android has \percMarketShare\ of the smartphone market and while the core development is controlled by Google, we know of \daNumManufacturers\footnote{\label{foot:dadata}We computed this from the \da~\cite{Wagner2013} data, see \S\ref{sec:android_update_process}.} manufacturers which make devices which run Android.
Android receives regular updates  which add new features and fix bugs, including security vulnerabilities.
Many of these manufacturers customise the version of Android they ship and sometimes network operators (of which we know of \daNumOperators\dafoot) make further modifications.
Hence when Google produces an update to Android, the update may have to pass through the manufacturer and operator before reaching the user.
The manufacturer has little financial incentive to perform this work as they have already sold the device.
Operators are cautious because if they ship a broken update then they will get many support requests which is expensive.
However there is ongoing legal action to force operators to ship updates for security vulnerabilities~\cite{Soghoian2013}.\todolater{Check on the status of this legal action}
Many phones are sold on long (12--24 month) contracts with monthly payments yet stop receiving updates before the end of the contract,\todolater{Is there a survey of the distribution of contract lengths and phone purchase methods?} whereas Windows XP could be purchased for a one-off payment in October 2001 and received security updates until April 2014.
Corporate or public sector buyers are encouraged to purchase secure devices, but we have found little concrete guidance. For example, CESG, which advises the UK government on how to secure its computer systems, recommends picking Android devices from manufactures which are good at shipping security updates promptly~\cite{CESG2013} but it does not state which manufacturers are better.\todolater{Which other advisory organisations give that advice?}
We provide this information in~\S\ref{sec:security_scoring}.

Android relies on the Linux kernel to provide compartmentalisation and runs each app as a separate Unix `user'.
The Linux kernel is a large piece of software and so there will likely be new root exploits found in it~\cite{TODO}.
Manufacturers of Android devices have also accidentally added further root exploits when customising it~\cite{Grace2012}.
Between 36.7\%~\cite{Zhou2012b} and 40\%~\cite{Zhou2012a} of malware for Android contained root exploits in 2012.
While this shows that much Android malware does not need a root exploit to work, a significant proportion does include root exploits.
This malware is also more dangerous because while ordinary malware can be remotely removed by Google through the Play store, once a root exploit has been used this is no longer guaranteed to work.

The proportion of Android devices in the \da\ data which were running versions of Android known to have root vulnerabilities at the time is shown in Figure~\ref{fig:proportioninsecure}.
It shows the proportion of devices which are definitely vulnerable as their build was released before the vulnerability was discovered, those which might be insecure but could contain a backported fix and those which are secure as the version of Android they were running contained no known vulnerabilities on that date.

\begin{figure}[!b]
\centering
\includegraphics[width=\columnwidth]{figures/proportioninsecure}
%TODO make this graph shorter
\caption{Proportion of devices running insecure, maybe insecure and secure versions of Android against time}
\label{fig:proportioninsecure}
\end{figure}

In the past users of Microsoft IIS were advised to switch to a different web server due to the frequent discovery of vulnerabilities~\cite{Pescatore2001}.
This pressure encouraged Microsoft to improve the security of their products~\cite{TODO}.
Part of the reason why deployed devices do not receive updates in a timely manner is that there is little data on what is actually happening and so it is not possible to determine which manufacturers or operators are doing a better job.
Hence there is little incentive for manufacturers and operators to do a better job.
We provide this data  in~\S\ref{sec:security_scoring}.

The contributions of this paper are:
\begin{itemize}
 \item We characterise the Android update ecosystem, showing how updates flow between entities, and quantify vulnerabilities by determining what proportion of devices they affected and how that changes over time.
 \item We collate information on vulnerabilities which affect Android and investigate their time-lines.
 \item We measure the security of Android according to three metrics and compare different manufacturers and device models to allow device purchasers to differentiate between them based on security.
 \todo{arb33: make this more concrete}
 \todo{arb33: Do we want \avo\ to provide our latest recommendations -- legal advice required}
\end{itemize}

>>>>

\section{Introduction}
%Android is the most popular smartphone operating system today with \percMarketShare\ marketshare.

Support for third-party apps is a key feature of modern smartphone operating systems.
Such apps are written by many developers from a wide range of backgrounds, which means neither operating system vendors nor users can, in general, fully trust app developers.
Therefore, in order to secure personal data and prevent the use of expensive operations such as the sending of premium-rate text messages, operating system vendors provide a protected execution environment, or \emph{sandbox}, for apps.

In this paper we quantify the extent to which the app sandbox provided by Android protects devices from malicious apps attempting to gain control over the entire handset.
%We have analysed Android because it is the most popular smartphone operating system by market share (~\percMarketShare) and 
Previous research has shown that in 2012, between 36.7\%~\cite{Zhou2012b} and 40\%~\cite{Zhou2012a} of malware for Android contained exploits designed to enable a malicious app to break the sandbox.
What is not known however is the proportion of Android devices which are vulnerable to these exploits over time.
To the best of our knowledge, this is the first paper to provide detailed numbers.

Not all malicious apps need to break the sandbox in order to misbehave.
Indeed, the above figures suggest the majority of malicious apps do not attempt to do so.
We focus on this issue because malware which is able to take full control of a handset can do significantly more harm and is much harder to remove.
Ordinary malware can uninstalled by the user or remotely by Google through the Play Store, but such removal is unlikely to work once the security features of Android are broken.

The development of Android is distributed among many parties.
The security of Android relies on many open source projects, include the Linux kernel, OpenSSL and BouncyCastle as well as Google who build the core platform. 
In addition, manufacturers (of which we are aware of \daNumManufacturers\footnote{\label{foot:dadata}We computed this from the \da~\cite{Wagner2013} data, see \S\ref{sec:android_update_process}.}) and network operators (\daNumOperators\dafoot) may make further modifications before devices are shipped to customers. 
Understanding this ecosystem is important as manufacturers have introduced additional vulnerabilities in the past~\cite{Grace2012}. 
We present a better understanding of both the ecosystem of Android development and associated vulnerabilities more fully in \S~X.

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figures/proportioninsecure}
\caption{Proportion of devices running insecure, maybe insecure and secure versions of Android against time}
\label{fig:proportioninsecure}
\end{figure}

In order to determine whether an individual smartphone is vulnerable to a particular exploit, we need to combine vulnerability data with the operating system version string and additional update information from the handset. 
For this we use handset data gathered from over 18,000 participants in the \da~\cite{} project.
This analysis allows us to plot the proportion of Android devices in the \da\ data which were running versions of Android known to have vulnerabilities at the time and is shown in Figure~\ref{fig:proportioninsecure}.
The Figure shows the proportion of devices split into three categories: 
(1) \emph{secure} because they are running a version of Android with no known vulnerabilities; 
(2) \emph{insecure} because the device was running an insecure version of Android with a build released before the vulnerability was fixed or discovered; and 
(3) \emph{maybe secure} because the device is running an insecure version of Android but it received an update which we did not see until after the vulnerability was fixed and therefore this update may contain a backport of the fix.
Further analysis into the propagation of updates in Android are presented in \S~Y.

It is clear from Figure~\ref{fig:proportioninsecure} that the security of the Android platform is far from perfect.
One solution is regulation, and indeed there is ongoing legal action to force operators to ship updates for security vulnerabilities~\cite{Soghoian2013}.\todolater{Check on the status of this legal action}
Indeed, many smartphones are sold on 12--24 month contracts, and yet our data shows many devices do not receive many security updates. 
In contrast, Windows XP could be purchased for a one-off payment in October 2001 and received security updates until April 2014.

An alternative to regulation is better public data. 
Hard numbers provide an incentive to both manufacturers and operators to provide updates.
Currently corporate and public sector buyers are encouraged to purchase secure devices, but we have found little concrete guidance on the specific makes and models which provide the best security. 
For example, CESG, which advises the UK government on how to secure its computer systems, recommends picking Android devices from manufactures which are good at shipping security updates promptly~\cite{CESG2013} but it does not state which manufacturers are better.
To address this, we developed a scoring system and provide numbers on the historic performance of devices in the \da\ project in~\S\ref{sec:security_scoring}.

%REMOVED BY ARB
%The manufacturer has little financial incentive to perform this work as they have already sold the device.
%Operators are cautious because if they ship a broken update then they will get many support requests which is expensive.
%
%In the past users of Microsoft IIS were advised to switch to a different web server due to the frequent discovery of vulnerabilities~\cite{Pescatore2001}.
%This pressure encouraged Microsoft to improve the security of their products~\cite{TODO}.

In summary, the contributions of this paper are:
\begin{itemize}
 \item We characterise the Android update ecosystem, showing how updates flow between entities, and quantify vulnerabilities by determining what proportion of devices they affected and how that changes over time.
 \item We collate quantitive information on vulnerabilities which affect Android and investigate their time-lines.
 \item We measure the security of Android according to three metrics and compare different manufacturers and device models to allow device purchasers to differentiate between them based on security.
\end{itemize}
